### 2018-03-15 paper/3-abstractions.tex
Add some text and TODOs for S3.6-S3.7

Despite limitations, \hs{busy} can easily handle the example \hs{sprsh1}
from the previous subsection~\S\ref{sec-task}. Below we initialise the store with
\cmd{A1} set to 10 and all other cells set to 20.

@@

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of executing the
    \hs{task} matches the value in the \hs{ideal} store:
    \vspace{-1mm}
    \[
    \hs{execute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
    \vspace{-1mm}
    \[
        \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
    \]
    In other words, we require that the output \hs{key} has the correct value,
    but do not impose any restrictions on intermediate keys, therefore
    permitting shallow cloud builds.
    \item The \hs{store}, \hs{result} and \hs{ideal} all agree on the inputs
    that belong to the transitive closure of \hs{key}'s dependencies:
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{key}~\hs{]}
    \]
    \vspace{-6mm}
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{key}~\hs{]}
    \]
    where the implementation of \hs{inputs} will be provided in the next
    subsection~\S\ref{sec-deps}, which is dedicated to the task dependency
    analysis. This requirement guarantees that (i) no inputs were corrupted
    during the build, and (ii)~the \hs{ideal} store has constraints not only
    on the output, but also on the inputs.
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{-2mm}
\rule{\textwidth}{0.4pt}

@@

The \hs{Task} type and the example above might seem mysterious. Why insist on
\hs{Task} to be polymorphic in \hs{f}? Why choose the \hs{Applicative}
constraint?

The answer to both questions is that we can statically analyse an applicative
compute and extract all dependencies of a key using the following function:

@@

\noindent
We can use this function to extract dependencies of \hs{add}, as shown in the
following GHCi session:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies add "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies add "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

\noindent
It is important to note that we extract the dependencies without providing any
values to the compute. This guarantees that no actual computation is performed
during the analysis -- we cannot execute the function \hs{(+)} if we have no
values for it! Static analysis of applicative computations is not new and has
been used to similar effect before, e.g. see~\cite{free-applicatives}.

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store key = if key == "A1" then 10 else 20
@\ghci@ execute add store "A1"
Nothing
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ execute add store "B1"
Just 30
\end{minted}

\subsection{Left-over stuff, need to relocate}

We cannot statically analyse \hs{select} using the function \hs{dependencies}
from the previous section, but we can still \hs{execute} it on a given store:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store key = if key == "A1" then 10 else 20
@\ghci@ execute select store "B1"
Just 20
\end{minted}
\vspace{1mm}

\noindent
In the example above $\cmd{C1} \neq 1$, hence the \hs{else} branch is taken.
To extract the dependencies of \hs{select} we introduce the function \hs{track}:
a combination of \hs{execute} and \hs{dependencies} that takes the store~\store
as input and computes both the resulting value and the list of its dependencies:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
track :: Task Monad k v -> (k -> v) -> k -> Maybe (v, [k])
track task store = fmap runWriter . task (\k -> writer (store k, [k]))
\end{minted}
\vspace{1mm}

\noindent
The implementation uses the \hs{Writer} monad (\S\ref{sec-appendix-transformers}):
we feed \hs{fetch k = writer (store k, [k])} to the compute, which in addition
to fetching a value from the store, tracks the corresponding key. We can
\hs{track} both applicative and monadic compute:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store key = if key == "A1" then 10 else 20
@\ghci@ track add store "B1"
Just (30,["A1","A2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ track select store "B1"
Just (20,["C1","A2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store' key = if key == "A1" then 10 else 1
@\ghci@ track select store' "B1"
Just (10,["C1","A1"])
\end{minted}
\vspace{1mm}

\noindent
As expected, the dependencies of \hs{select} are determined by the contents of
the store.

>>>>

Despite limitations, \hs{busy} can easily handle the example \hs{sprsh1}
from the previous subsection~\S\ref{sec-task}. In the GHCi session below we
initialise the store with \cmd{A1} set to 10 and all other cells set to 20.

@@

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of executing the
    \hs{task} matches the value in the \hs{ideal} store:
    \vspace{-1mm}
    \[
    \hs{execute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
    \vspace{-1mm}
    \[
        \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
    \]
    In other words, we require that the output \hs{key} has the correct value,
    but do not impose any restrictions on intermediate keys, therefore
    permitting shallow cloud builds.
    \item The \hs{result}, \hs{store} and \hs{ideal} all agree on the inputs
    that belong to the transitive closure of \hs{key}'s dependencies with
    respect to the \hs{result}:
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{store}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    \vspace{-6mm}
    \[
        \hs{and}~\hs{[}~\hs{getValue}~\hs{result}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}~\hs{|}~\hs{k}~\hs{<-}~\hs{inputs}~\hs{task}~\hs{result}~\hs{key}~\hs{]}
    \]
    where the implementation of \hs{inputs} will be provided in the next
    subsection~\S\ref{sec-deps}, which is dedicated to the task dependency
    analysis. This requirement guarantees that (i) no inputs were corrupted
    during the build, and (ii)~the \hs{ideal} store has constraints not only
    on the output, but also on the inputs.
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{-2mm}
\rule{\textwidth}{0.4pt}

@@

\todo{SLPJ}{Please edit as you see fit. We need \hs{dependencies} and
\hs{track} to be defined for future sections. We also promised to define
\hs{inputs} in the previous section, but I think it's not too critical and can
be skipped/moved to Appendix if need be.}

...

We can statically analyse an applicative task and extract all dependencies of a
key using the following function:

@@

\noindent
We can use this function to extract dependencies of \hs{sprsh1}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "A1"
[@@]
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies sprsh1 "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

\noindent
It is important to note that we extract the dependencies without supplying any
values to the build task. This guarantees that no actual computation is performed
during the analysis -- we cannot execute the function \hs{(+)} if we have no
values for it! Static analysis of applicative computations is not new and has
been used to similar effect before, e.g. see~\cite{free-applicatives}.

@@

To extract the dependencies of monadic tasks we introduce the function \hs{track}:
a combination of \hs{execute} and \hs{dependencies} that takes a store as input
and computes both the resulting value and the list of its dependencies:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
import Control.Monad.Writer
\end{minted}
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
track :: Task Monad k v -> Store i k v -> k -> Maybe (v, [k])
track task store = fmap runWriter . task (\k -> writer (getValue store k, [k]))
\end{minted}
\vspace{1mm}

The implementation uses the \hs{Writer} monad (\S\ref{sec-appendix-transformers}):
we feed \hs{fetch k = writer (getValue store k, [k])} to the task, which in
addition to fetching a value from the store, tracks the corresponding key. We
can \hs{track} both applicative and monadic tasks:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store = initialise () (\key -> if key == "A1" then 10 else 20)
@\ghci@ track sprsh1 store "B1"
Just (30,["A1","A2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ track sprsh2 store "B1"
Just (20,["C1","A2"])
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store' = initialise () (\key -> if key == "A1" then 10 else 1)
@\ghci@ track sprsh2 store' "B1"
Just (10,["C1","A1"])
\end{minted}
\vspace{1mm}

\noindent
As expected, the dependencies of \hs{sprsh2} are determined by the contents of
the store.

...

Here is a draft implementation of \hs{inputs} used in the definition of
build system correctness in \S\ref{sec-build-correctness}:

\begin{minted}[xleftmargin=10pt]{haskell}
inputs :: Eq k => Task Monad k v -> Store i k v -> k -> [k]
inputs task store key = filter (isInput task) (closure deps key)
  where
    deps k = maybe [] snd (track task store k)

closure :: Eq a => (a -> [a]) -> a -> [a] -- Standard graph transitive closure

data Proxy a = Proxy

isInput :: Task Monad k v -> k -> Bool
isInput task = isNothing . task (const Proxy)
\end{minted}

\subsection{Free tasks (?)}

\todo{AM}{Introduce free constructions \hs{TaskA} and \hs{TaskM} giving
intuitions on what tasks really are. Requires some further discussion with Neil.
This will take at least one page.}
